Value-Suppressing Uncertainty Maps

### 2017-12-06 paper/main.tex
Some bibliography fixes.

\section{Value-Suppressing Uncertainty Palettes}

%TODO need some example figure here, showing tree structure perhaps?
%\flowFig

Value-Suppressing Uncertainty Palettes (VSUPs) are a technique for creating bivariate maps of data \emph{value} and \emph{uncertainty}. VSUPs make two central assumptions about bivariate maps:

@@

JavaScript code for generating VSUPs for use in D3~\cite{bostock2011d3} charts is available at \url{URL-REMOVED-FOR-REVIEW}.

@@

When selecting a visual channel to represent uncertainty, we recommend channels with both a strong semiotic connection to uncertainty (such as those recommended by MacEachren et al.~\cite{maceachren2012visual}) as well as a relatively large number of perceptually distinguishable levels (such as those evaluated by Boukhelifa et al. \cite{boukhelifa2012evaluating}). Here, we focus on a combination of increasing luminance and decreasing saturation. Correll \& Gleicher~\cite{correll2013error} provide evidence that even audiences without statistical training can successfully interpret uncertainty information encoded in these channels.

@@

\subsection{Viral Mutation}
%TODO update this figure
\viralFig

Next generation sequencing allows biologists to collect and analyze large amounts of genomics data. However, these techniques often create data quality issues, as they must align large numbers of potentially ambiguous ``reads'' of relatively small numbers of base pairs. Using a dataset from a group of virologists interested in the population dynamics of viral mutation~\cite{o2012conditional}, we visualized the variability in viral genomes across 12 populations of the simian immunodeficiency virus (SIV). In the dataset here, it is important not only to identify hotspots in the viral genome (locations with high rates of mutation or variability), but also to not be distracted by false positives. The traditional bivariate map (\figref{fig:viral2d}) has many locations with apparent inter-population variance.

VSUPs (\figref{fig:viralVsum}) enable biologists to explore interesting patterns while deemphasizing noise. Orange hot spots towards the beginning (the second column) and end of the genome (around reference nucleotide $9,500$) could be considered potential regions of interest in the 2D map; in the VSUP these are encoded as highly uncertain and of ambiguous value. In fact, these regions have systematically low coverage; the lack of reliable data for these regions should discourage analysts from giving their apparent pattern equal consideration from the stronger signals elsewhere in the genome.

\section{Evaluation}
%TODO update conditions fig
\conditionFig

@@

Experimental materials, including data tables and stimuli generation code, are available at \url{URL-REMOVED-FOR-REVIEW}.

>>>>

\section{Value-Suppressing Uncertainty Palettes}
Value-Suppressing Uncertainty Palettes (VSUPs) are a technique for creating bivariate maps of data \emph{value} and \emph{uncertainty}. VSUPs make two central assumptions about bivariate maps:

@@

JavaScript code for generating VSUPs for use in D3~\cite{bostock2011d3} charts is available at \url{https://github.com/uwdata/spatial-sampling}.

@@

When selecting a visual channel to represent uncertainty, we recommend channels with both a strong semiotic connection to uncertainty (such as those recommended by MacEachren et al.~\cite{maceachren2012visual}) as well as a relatively large number of perceptually distinguishable levels (such as those evaluated by Boukhelifa et al. \cite{boukhelifa2012evaluating}). Here, we focus on a combination of increasing luminance and decreasing saturation. Correll \& Gleicher~\cite{correll2014error} provide evidence that even audiences without statistical training can successfully interpret uncertainty information encoded in these channels.

@@

%\subsection{Viral Mutation}
%\viralFig

%Next generation sequencing allows biologists to collect and analyze large amounts of genomics data. However, these techniques often create data quality issues, as they must align large numbers of potentially ambiguous ``reads'' of relatively small numbers of base pairs. Using a dataset from a group of virologists interested in the population dynamics of viral mutation~\cite{o2012conditional}, we visualized the variability in viral genomes across 12 populations of the simian immunodeficiency virus (SIV). In the dataset here, it is important not only to identify hotspots in the viral genome (locations with high rates of mutation or variability), but also to not be distracted by false positives. The traditional bivariate map (\figref{fig:viral2d}) has many locations with apparent inter-population variance.

%VSUPs (\figref{fig:viralVsum}) enable biologists to explore interesting patterns while deemphasizing noise. Orange hot spots towards the beginning (the second column) and end of the genome (around reference nucleotide $9,500$) could be considered potential regions of interest in the 2D map; in the VSUP these are encoded as highly uncertain and of ambiguous value. In fact, these regions have systematically low coverage; the lack of reliable data for these regions should discourage analysts from giving their apparent pattern equal consideration from the stronger signals elsewhere in the genome.

\subsection{2016 Pre-Election Polling}

Using a public data set from Wikipedia~\cite{wiki}, we collected the statewide polls that were conducted closest to the 2016 U.S. presidential election, we created a choropleth map showing which candidate was leading in the polls. We calculated uncertainty as the size of the lead in terms of margin of error (lead/margin). The AP style guide for reporters suggests that a candidate leading by more than two margins of error is said to be ``significantly ahead'' in the polls~\cite{french1988associated}. As the lead becomes closer to the margin of error, there is more doubt about the electoral outcome.

By providing more color categories when the margin of error was relatively small, the VSUP map shows that, in many of the states leaning towards Hillary Clinton, her lead was quite small. The VSUP map, by limiting inferences about which candidate was leading when the uncertainty is high, also prevents the viewer from making strong claims about the outcome of the election, even if a state is leaning one way or the other. Given the relative volatility in polling data, this caution was warranted. An additional benefit to the VSUP map is that, as the diverging blue-red color map passes through a central whitepoint, it can be difficult to distinguish between uncertainty levels (which also interpolates to white) in the 2D map. The VSUP's tree-like structure separates these problematic colors.

\section{Evaluation}
\conditionFig

@@

Experimental materials, including data tables and stimuli generation code, are available at \url{https://github.com/uwdata/spatial-sampling}.
